---
title: "Duration of and time to response in oncology clinical trials from the perspective of the estimand framework"
subtitle: "Code accompanying the paper: Case Study in Mantel Cell Lymphoma (Section 4)"
author: "Hans-Jochen Weber"
date: 'Last modification: `r Sys.setlocale("LC_TIME", "C"); format(Sys.Date(), "%d %b %Y")`'
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    number-sections: true
    number-depth: 2
bibliography: stat.bib
---

# Background

Code for the paper @weber_23 written by the [duration of response task force](https://oncoestimand.github.io/oncowg_webpage/docs/taskforces.html#duration-of-responses) of the [oncology estimand WG](http://www.oncoestimand.org).

# Purpose of this document

This R markdown file provides easy accessible code to reproduce computations in the paper. The github repository where this document is hosted is available [here](https://github.com/oncoestimand/dor).

# Setup

## Packages

```{r, setup, warning = FALSE, message = FALSE}
# --------------------------------------------------------------
# Packages
# --------------------------------------------------------------
packages <- c("gtsummary", "survival", "cmprsk", "survminer", "msm", 
              "mstate", "tidyverse", "gt", "survRM2", "PBIR", 
              "labelled", "swimplot", "ggplot2")

for (i in 1:length(packages)){library(packages[i], character.only = TRUE)}
```

## 2.2 Get data

The source data provide the response status at time points $C_2, \ldots, C_{22}$. The data set has no deaths. For assessing time to response (TTR), duration of response (DOR), or overall response rate (ORR) deaths are handled in the same way as progression. The dataset also has no data on intercurrent events such as e.g. start of new therapy. Thus, we assume none occurred. A treatment cycle has 28 days.

```{r, data, warning = FALSE, message = FALSE}
# --------------------------------------------------------------
# Data
# --------------------------------------------------------------
dat <- read_csv("dor.csv", show_col_types = FALSE)
gt(dat) %>% 
  tab_header(title = "Listing: Responses by cycle") %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_title()) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_column_labels())
```

## Data preparation

```{r, data_prep}
# --------------------------------------------------------------
# Data preparation
# --------------------------------------------------------------

# Don't need Patient ID column
data <- dat[, -1]

# Time of assessment in Cycles, Days, Months
cycles <- as.numeric(sub('.', '', colnames(data))) 

# cycle duration in days
cycle.duration <- 28 

days <- function(x){
  x * cycle.duration
}
days.per.month <- 365.25 / 12
months <- function(x){
  days(x) / days.per.month
}

n <- dim(data)[1]
```

## Endpoint derivations

Endpoint derivations below work for 

- complete data,
- with no missing assessments prior to last assessment, 
- no assessments after progression and 
- all patients have at least 1 assessment before drop-out/death. 

This is the case with the data set in the case study.

```{r, endpoint_derivation}
# --------------------------------------------------------------
# Endpoint derivations
# --------------------------------------------------------------

# Pick the best response by patient 
BOR <- function(x){
  # x: data matrix
  if(any(x == "CR", na.rm = TRUE)){out <- "CR"}
  else{
    if(any(x == "PR", na.rm = TRUE)){out <- "PR"}
    else{
      if(any(x == "SD", na.rm = TRUE)){out <- "SD"}
      else{
        out<-"PD"
      }
    }
  }
}

bor <- apply(data, 1, BOR)

# Best overall response is CR or PR
OR <- function(x){
  # x: BOR vector
  ifelse(x == "CR" | x == "PR", TRUE, FALSE)
}

or <- OR(bor)

# Any assessment with PD?
Any.PD <- function(x){
  # x: data matrix
  ifelse(any(x == "PD",na.rm = TRUE), TRUE, FALSE)
}

any.pd <- apply(data, 1, Any.PD)

# OR and an assessment with PD?
or.pd <- ifelse(or == TRUE & any.pd, TRUE, FALSE)

# No OR but an assessment with PD? 
noor.pd <- ifelse(or == FALSE & any.pd, TRUE, FALSE)

# No OR or an assessment with PD? 
noor.orpd <- ifelse(or == FALSE | any.pd, TRUE, FALSE)

# Last assessment in study (cycle number)
last <- numeric(n)
for(i in 1:n){
  last[i] <- cycles[length(data[i, which(!is.na(data[i, ]))]) ]   
} 

# TTR: Time to response. 
# - For patients with OR use date of first assessment 
# with OR (response event) 
# - For patients with no OR nor PD use last assessment 
# (censored, could get response if further assessments taken). 
# - For patients with no OR and a PD use last assessment 
# (censored, competing risk PD).

ttr <- numeric(n)
for(i in 1:n){
  if(or[i]){
    ttr[i] <- cycles[min(which(data[i, ] %in% c("CR", "PR")))] }
  else{
    ttr[i] <- last[i]}
}

# DOR: Duration of response
# For patients with no response: 0
# For patients with response and no PD: Date of last assessment - 
# Date of response (censored).
# For patients with response and PD: 
# Date of last assessment - Date of response (event) 

dor <- ifelse(or, last - ttr, 0)

# For time in response: status
tir.status <- ifelse(noor.orpd, 1, 0)
tir.status <- ifelse(or.pd, 1, tir.status)

# For competing risks: TTR status 1 = OR, 2 = no OR but PD, 
# 0 = no OR, no PD
ttr.status <- ifelse(or, 1, 0)
ttr.status[!or & noor.pd] <- 2

# TTP: Time to progression (equivalent to PFS in this dataset)
# For patients with PD based on PD assessment date (= event)
# For patients without PD based on last assessment date (= censored)
# Note that in this data set, if PD occurs, it is at the 
#   last assessment date.
# Note that in this data set there are no deaths.
# Therefore in this data set, ttp = pfs = last assessment

ttp.status <- ifelse(any.pd, 1, 0)
ttp <- last

# Use last observed follow-up as date of TTR if progression 
# (Table 3, estimand 2)
sttr <- ifelse(any.pd == 1 & or == 0, max(ttp), ttr)

# TTP censored at response
# For patients with OR: response date
# For all other patients: last
ttp.cens <- ifelse(or, ttr, last)

data.ext <- cbind(dat, bor, or, any.pd, or.pd, noor.pd, 
                  noor.orpd, last, ttr, sttr, dor, ttp, 
                  ttp.status, ttp.cens, tir.status)
```

# BOR and ORR

Consider any best response of PR or CR. 

```{r, BOR, message = FALSE}
# --------------------------------------------------------------
# Best overall response
# --------------------------------------------------------------
rr <- data.ext %>% 
      select(or)
r <- rr %>% 
     tbl_summary(label = or ~ "Best overall response",
                 digits = list(all_categorical() ~ c(0, 1)))
as_gt(r)
```

# Duration of response 

## Conditional cDOR (KM analysis, responders only)

```{r, cDOR}
# --------------------------------------------------------------
# Conditional duration of response
# --------------------------------------------------------------

# responders only
data.lim <- subset(data.ext, or == TRUE) 

# or.pd = event indicator, TRUE (1) = event
fit_cDOR <- survfit(Surv(months(dor), or.pd) ~ 1, data = data.lim) 

# Figure 2
ggsurvplot(fit_cDOR, data = data.lim, risk.table = TRUE, conf.int = F,
           title = "Conditional duration of Response (responders only)", 
           submain = "Kaplan-Meier estimates", 
           break.x.by = 3,
           ylab = "Prob(progression or death)",
           xlab = "Months from start of treatment",
           legend.title = "")
```

```{r, cDOR_KM_estimates, message = FALSE}
# Table 4, row 1
as_gt(tbl_survfit(fit_cDOR, times = months(c(6, 9, 12))))
```

## Time in response (EMA definition, all patients)

```{r, Time_in_response}
# --------------------------------------------------------------
# Time in response (all patients): EMA definition
# --------------------------------------------------------------

# tir.status = event indicator, TRUE (1) = event
fit_DOR <- survfit(Surv(months(dor), tir.status) ~ 1, data = data.ext) 

ggsurvplot(fit_DOR, data = data.ext, risk.table = TRUE, conf.int = F,
           title = "Time in Response (EMA definition, all patients)", 
           submain = "Kaplan-Meier estimates", 
           break.x.by = 3,
           xlab = "Months from start of treatment",
           legend.title = "")
```

```{r, Time_in_response_KM_estimates, message = FALSE}
# Table 4, row 2
as_gt(tbl_survfit(fit_DOR, times = months(c(6, 9, 12))))
```

## Probability of being in response (all patients)

```{r, PBIR}
# --------------------------------------------------------------
# Probability of being in response [months] (PBIR package)
# --------------------------------------------------------------
fit_PBIR<-PBIR1(t2PROGRESSION = months(data.ext$ttp),
                STATUS_PROGRESSION = data.ext$any.pd,
                t2RESPONSE = months(data.ext$ttr),
                STATUS_RESPONSE = data.ext$or,
                time = months(c(1:10)))

# Table 4, row 3
fit_PBIR %>% 
  gt()  %>% 
  fmt_number(decimals = 2, columns = c(time, PBIR, std, ci.low, ci.up)) %>%
   tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_column_labels())
```

```{r, PBIR_plot}
fit_PBIR_pl<-PBIR1(t2PROGRESSION = months(data.ext$ttp),
                   STATUS_PROGRESSION = data.ext$or.pd ,
                   t2RESPONSE = months(data.ext$ttr),
                   STATUS_RESPONSE = data.ext$or)

tt   <- fit_PBIR_pl$time
diff <- fit_PBIR_pl$PBIR
B    <- length(tt) + 1
tt   <- c(0, tt)
diff <- c(0, diff)

tt   <- rep(tt, rep(2, B))[-1]
diff <- rep(diff, rep(2, B))[-(2 * B)]

plot(range(c(0, max(tt))), range(c(0, 1)), 
     xlab = "Months from start of treatment", ylab = "PBIR",
     lwd = 2, type = "n", main = "Probability of being in response (PIBR)")
lines(tt, diff, lwd = 2, col = 3)
```

## Mean duration of response (all patients)

```{r, mean_DOR, message = FALSE, warning = FALSE}
# --------------------------------------------------------------
# Mean duration of response [months] (PBIR package)
# --------------------------------------------------------------

# Table 4, row 4
mduration(t2PROGRESSION = months(data.ext$ttp),
          STATUS_PROGRESSION = data.ext$any.pd,
          t2RESPONSE = months(data.ext$ttr), 
          STATUS_RESPONSE = data.ext$or)
```

# Time to Response

## cTTR (responders only, KM analysis)

```{r, cTTR_KM, message = FALSE, warning = FALSE}
# --------------------------------------------------------------
# Conditional TTR
# --------------------------------------------------------------
data.res <- subset(data.ext, data.ext$or == 1)

# or = event indicator, TRUE (1) = event
fit_cTTR <- survfit(Surv(months(data.res$ttr), data.res$or) ~ 1, 
                    data = data.res) 

# Table 5, row 5
surv_median(fit_cTTR)

# Table 5, row 6
surv_summary(fit_cTTR)

# KM plot cTTR (Figure 3)
ggsurvplot(fit_cTTR, data = data.res, 
           risk.table = TRUE,
           conf.int = FALSE,
           surv.median.line = "h",
           main = "Time to Response", submain = "Kaplan-Meier estimates",
           xlim = c(0, 10),
           break.x.by = 1, ylab = "Prob(achieving response)",
           xlab = "Months from start of treatment",
           legend.title = "")  
```

## cTTR censoring at last patient last visit (KM analysis)

For this analysis we set the censoring date to the maximum follow-up date observed in the study for patients with progression prior to achieving response. 

```{r, TTR_censor_last_visit, message = FALSE, warning = FALSE}
# --------------------------------------------------------------
# Alternative time to response 
# --------------------------------------------------------------

# or = event indicator, TRUE (1) = event
fit_cTTR_TP <- survfit(Surv(months(sttr), or) ~ 1, data = data.ext) 

ggsurvplot(fit_cTTR_TP, data = data, risk.table = TRUE,
           title = "Time to Response (censor at last visit, responders only)",
           conf.int = FALSE, submain = "Kaplan-Meier estimates",
           xlab = "Months from start of treatment",
           legend.title = "")

# Table 5, row 7
surv_median(fit_cTTR_TP)

# Table 5, row 8
surv_summary(fit_cTTR_TP)
```

## TTR (while-progression free and alive, all patients, competing risks analysis)

Cumulative incidence function (months). In the figure below, event = 1 represents response, event = 2 represents PD without response. We derive the estimate at Month 6 with confidence interval.

```{r, CIR, message = FALSE, warning = FALSE}
# --------------------------------------------------------------
# Cumulative incidence function (cmprsk package)
# --------------------------------------------------------------

# Table 5, row 9
fit_CIF <- cmprsk::cuminc(ftime = months(ttr), fstatus = ttr.status)

z6 <- timepoints(fit_CIF, times = c(6))
z6cil <- z6$est - sqrt(z6$var) * (-qnorm(.025))
z6ciu <- z6$est + sqrt(z6$var) * (-qnorm(.025))
c("response (95% CI):", 1 - round(c(z6$est[1], z6cil[1], z6ciu[1]), 3))

# CIF plot
cif <- ggcompetingrisks(fit_CIF, conf.int = FALSE, multiple_panels = FALSE,
                 title = "Time to response (CIF, all patients)", 
                 palette = "jco", xlab = "Months from start of treatment")
cif2 <- cif + scale_color_manual(name = "Event", values = c(1, 2),
                                labels = c("Response", 
                                           "Progression without response"))
ggpar(cif2, xlim = c(0, 9.5), xticks.by = 1)
```

# 6 Swimmer plot

```{r, swimmer_plot, message = FALSE, warning = FALSE}
# --------------------------------------------------------------
# Swimmer plot (swimplot package)
# --------------------------------------------------------------

# Figure 4
# Response continues at study end (no progression)
data.ext$cont <- ifelse(data.ext$any.pd == FALSE, data.ext$ttp, NA)

# Progression at study end
data.ext$prog <- ifelse(data.ext$any.pd == TRUE, data.ext$ttp, NA)
data.ext$Progression <- ifelse(data.ext$any.pd == TRUE, "Progression", NA)

follow_p <- swimmer_plot(df = data.ext, id = 'ID', end = 'last', width = .85, 
                         name_fill = 'bor')

resp <- follow_p + swimmer_lines(df_lines = data.ext, id = 'ID', 
                                 start = 'ttr', end = 'ttp', size = 1, 
                                 col = c("black"))  +
  
  swimmer_points(df_points=data.ext, id = 'ID', time = 'prog', size = 2.5,
                 fill = 'white', col = 'black', name_shape = "Progression") +
  
  scale_y_continuous(name = "Time since start of treatment (cycles)",
                     breaks = c(0, 2, 4, 6, 8, 10, 13, 16, 19, 22)) +
  
  swimmer_arrows(df_arrows = data.ext, id = 'ID', arrow_start = 'cont',
                 cont = 'any.pd', type = "open", cex = 1, name_col = 'bor') +
  
  scale_fill_manual(name = "Best overall response", 
                    values = c("CR" = "#4daf4a",
                               "PR" = "#377eb8",
                               "SD" = 'orange',
                               "PD" = '#e41a1c')) +
  
  scale_color_manual(name = "Best overall response", 
                    values = c("CR" = "#4daf4a",
                             "PR" = "#377eb8",
                             "SD" = 'orange',
                             "PD" = '#e41a1c')) +
  
  scale_shape_manual(name = "End of follow-up", values = c(Progression = 21),
                     breaks = c('Progression')) +
  
  annotate("text", x = 3.5, y = 28.45, 
           label = "Continued response", size = 3.25) +
  
  annotate("text", x = 2.5, y = 28.25, 
           label = sprintf('\u2192'), size = 8.25) +
  
  coord_flip(clip = 'off', ylim = c(0, 23))

resp
```

# References